---
title: "Indexing Temporal Relations for Range-Duration Queries"
subtitle: "Supplemental material"
author:
  - Matteo Ceccarello <mceccarello@unibz.it>
  - Anton Dignös <dignoes@inf.unibz.it>
  - Johann Gamper <gamper@inf.unibz.it>
  - Christina Khnaisser <christina.khnaisser@usherbrooke.ca>
execute:
  echo: false
reference-location: margin
citation-location: margin
format:
  html:
    self-contained: true
---

This document complements the experimental results of the paper. 
In particular, in this document we provide interactive plots to explore the behavior of different index structures on mixed query workloads.

```{css}
.plot {
  font-size: 12pt;
}
thead>tr {
  border-bottom-color: rgb(204, 204, 204);
  border-bottom-style: solid;
  border-bottom-width: 1px;
}
```

```{r setup, echo=F, message=F, warning=F}
library(tidyverse)
library(kableExtra)
library(units)
data <- read_csv("best.csv") %>%
  filter(
    dataset_name %in% c("UZ", "Flight", "Webkit", "MimicIII"),
    time_constraint %in% c("UZ", "UU", "-"),
    duration_constraint %in% c("U", "-")
  ) %>%
  select(
      dataset_name, time_constraint,
      duration_constraint, algorithm_name, qps
  ) %>%
  group_by(dataset_name, time_constraint, duration_constraint) %>%
  distinct(algorithm_name, qps) %>%
  replace_na(list(qps = 0)) %>%
  mutate(
      rank = dense_rank(desc(qps)),
      rank_str = str_c(" {\\footnotesize(", rank, ")}"),
      qps_num = qps,
      qps = qps %>% scales::number(accuracy=1)
  ) %>%
  select(
      dataset = dataset_name,
      time = time_constraint,
      duration = duration_constraint,
      algorithm_name,
      qps, qps_num
  ) %>%
  mutate(
      algorithm_name = case_when(
          algorithm_name == "BTree"          ~ "BTree",
          algorithm_name == "grid-file"      ~ "GridFile",
          algorithm_name == "period-index-*" ~ "Period-Index*",
          algorithm_name == "interval-tree"  ~ "ITree",
          algorithm_name == "rd-index-td"    ~ "RD-Index-td",
          algorithm_name == "rd-index-dt"    ~ "RD-Index-dt",
          algorithm_name == "RTree"          ~ "RTree",
          TRUE ~ algorithm_name
      ),
      algorithm_name = factor(algorithm_name,
          levels = c(
              "RD-Index-td",
              "RD-Index-dt",
              "GridFile",
              "Period-Index*",
              "RTree",
              "ITree",
              "BTree"
          ),
          ordered = TRUE
      )
  ) %>%
  mutate(
      query_type = case_when(
          time == "UZ" && duration == "U" ~ "qrd",
          time == "UU" && duration == "U" ~ "qrd",
          time == "-" && duration == "U"  ~ "qdo",
          time == "UZ" && duration == "-" ~ "qro",
          time == "UU" && duration == "-" ~ "qro",
          T ~ "unknown"
      ),
      query_type = factor(query_type, levels = c("qro", "qdo", "qrd"), ordered = T)
  ) %>%
  ungroup() %>%
  select(
      dataset,
      query = query_type,
      algorithm_name,
      qps, qps_num
  ) %>%
  arrange(algorithm_name)

# ojs_define(data = select(data, dataset, query, algorithm_name, qps_num) %>%
#   pivot_wider(names_from=query, values_from=qps_num))
# ojs_define(tradeoffs = read_csv("tradeoffs.csv"))
```

## Mixed-workload queries

The paper reports on the performance of various index structures for uniform workload queries, i.e. range-only queries, duration-only queries, or range-duration queries.

```{r}
#| tbl-cap: "Table 1, as reported in the paper"
#| tbl-cap-location: top
#| column: screen-inset

data %>% 
  select(dataset, query, algorithm_name, qps) %>%
  pivot_wider(
      names_from = "algorithm_name",
      values_from = "qps"
  ) %>%
  mutate(
      dataset = case_when(
          dataset == "UZ" ~ "Random",
          TRUE ~ dataset
      ),
      dataset = factor(dataset,
          levels = c("Random", "Flight", "Webkit", "MimicIII"),
          ordered = TRUE
      )
  ) %>%
  arrange(dataset, query) %>%
  kbl(align="r") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = T, font_size = 14)

```

A natural question is: how does the performance change when these three types of queries are mixed together?
Using data from the above table we can estimate performance figures for mixed workloads.
In particular, the inverse of the throughputs reported in the above table are the average running time of each specific type of query.
Taking the inverse of the weighted average (where weights are the fraction of queries of each type) is the throughput of the mixed workload:
$$
QPS_{mixed} =
\frac{1}{
  f_{ro} / QPS_{ro} +
  f_{do} / QPS_{do} +
  f_{rd} / QPS_{rd}
}
$$
where $f_{ro}$ is the fraction of range only queries in the workload,
$f_{do}$ is the fraction of duration only queries in the workload,
and $f_{rd}$ is the fraction of range-duration queries.

```{ojs}
// import { aq, op } from '@uwdata/arquero'
// import { ternaryPlot, barycentric } from "@julesblm/d3-ternary"
aq = require('arquero');
d3 = require("d3@7")
d3ternary = require("d3-ternary@2")
```

Using the following radio buttons, we can select on which dataset to focus on.

```{ojs}
viewof datasetName = Inputs.radio(["Random", "Flight", "Webkit", "MimicIII"], {value: "Random", label: "Dataset"})
```

Then, the ternary plot below allows to select different workloads by hovering the mouse on top of it.
Towards the top corner we have workloads composed for the most part by range only queries.
The bottom left corner is for workloads with a majority of duration only queries.
In the bottom right corner we have workloads with mainly range-duration queries.
The table above the plot reports the precise composition of the select workload.
At the right of the ternary plot, a bar chart displays the performance attained by the different index structures on the selected workload, in queries per second.
The ternary plot area's itself is colored with the color of the best performing index in each particular point.

```{ojs}
// Function to transform a triplet of number adding up to 1 into a triplet of integers adding up to 100
function roundTriplet(triplet) {
  const [x, y, z] = d3.map(triplet, x => Math.round(x * 100))
  const excess = x + y + z - 100;
  return [x, y, z - excess];
}
```

```{ojs}
// This is the table of all the performance and its tradeoffs
// table = aq.table(tradeoffs)
table = aq.loadCSV("tradeoffs.csv")
```

```{ojs}
// This is the map holding the best index for each workload
bestMapData = table.params({dataName: datasetName}).filter((d, $) => d.dataset == $.dataName)
  .derive({
    triplet: aq.escape(d => roundTriplet([d.frac_ro, d.frac_do, d.frac_rd]))
  })
  .groupby("dataset", "triplet")
  .orderby(d => -d.qps)
  .slice(0, 1)
```

```{ojs}
algoDomain = [
  "RD-index-td"   ,
  "RD-index-dt"   ,
  "Grid-File"     ,
  "Period-Index*" ,
  "R-Tree"        ,
  "Interval-Tree" ,
  "B-Tree"        
];
```

```{ojs}
// This is the subset of data that corresponds to the position under the mouse
triplet = (bestMap != 0)? roundTriplet(bestMap.coords) : [0,0,0]; // Return the impossible triplet if a selection has yet to be made
selectedSubset = d3.filter(table, 
  d => d.dataset == datasetName &&
    Math.round(d.frac_ro * 100) == triplet[0] && 
    Math.round(d.frac_do * 100) == triplet[1] && 
    Math.round(d.frac_rd * 100) == triplet[2])
```

```{ojs}
axisLabels = (g, labels) =>
  g
    .selectAll("text")
    .data(labels, d => d.label)
    .join(
      enter =>
        enter
          .append("text")
          .attr("dy", (d, i) => (i === 0 ? "2em" : ".5em"))
          .attr(
            "transform",
            (label, i) => `translate(${label.position})`
          )
          .attr("text-anchor", "middle")
          .text(label => label.label),
      update =>
        update.attr(
          "transform",
          (label, i) => `translate(${label.position})`
        )
    )
```

```{ojs}
valuesTable = (labels, plot) => html`
<table style="height:50px; width:100%">
  <thead>
    <tr style="width:33%">${labels.map((label) => `<th>${label}</th>`)}</tr>
  </thead>
  <tbody>
    <tr>${
      plot === 0
        ? `<td colspan="3">
            Hover your mouse over the triangle to select a workload
          </td>`
        : plot.coords.map(c => `<td>${d3.format(".0%")(c)}</td>`)
    } </tr>
  </tbody>
</table>
`
valuesTable(['Range only', 'Duration only', 'Range-duration'], bestMap)
```

::: {layout-ncol=2 .column-page}
```{ojs}
viewof bestMap = {
  const height = 350;
  const width = 500;
  const radius = Math.min(width, height) / 2;
  // to keep the triangle centered
  const yOffset = radius/4;
  // Set up the chart space
  const node = DOM.svg(width, height);
  const svg = d3.select(node);
  const chart = svg
    .append("g")
    .attr("transform", `translate(${width / 2} ${height / 2 + yOffset})`)
    .on("mousemove", handleMouseMove);

  const circle = chart.append("g");
  
  // Handler of mouse hover events. It stores the state
  // underlying the mouse in the SVG node itself.
  function handleMouseMove(d) {
    const xy = d3.pointer(d);
    const inverse = ternary.invert(xy);

    node.dispatchEvent(new CustomEvent("input"), { bubbles: true });
    node.value = {
      coords: inverse,
      data: d.target.__data__
    };
  }

  // Define transformer in ternary coordinates
  const normalBarycentric = d3ternary.barycentric()
    .a(d => d.frac_ro)
    .b(d => d.frac_do)
    .c(d => d.frac_rd);
  const ternary = d3ternary.ternaryPlot(normalBarycentric)
    .radius(radius)
    .labels(["Range only", "Duration only", "Range-Duration"]);

  const axisLabelsGroups = chart
    .append("g")
    .attr("class", "axis-labels")
    .append("g")
    .attr("class", "labels")
    .attr("font-size", 16);

  axisLabelsGroups.call(axisLabels, ternary.axisLabels());
  
  // Add the ternary components to the data
  const ternaryData = d3.map(bestMapData, (d) => {
    const [x, y] = ternary(d);
    return { x, y, ...d };
  });

  // border triangle
  const trianglePath = chart
    .append("path")
    .attr("d", ternary.triangle())
    .attr("fill", "transparent")
    .attr("stroke", "black")
    .attr("stroke-width", 6);

  // Set up color scale
  const color = d3.scaleOrdinal(d3.schemeTableau10)
    .domain(algoDomain);

  // Add the points
  chart.append("g")
    .selectAll("circle")
    .data(ternaryData)
    .join("circle")
    .attr("r", 2)
    .attr("cx", (d) => d.x)
    .attr("cy", (d) => d.y)
    .attr("fill", d => color(d.algorithm))
    .attr("stroke", "none");

  node.value = 0;

  return svg.node();
}
```

```{ojs}
Plot.plot({
  width: 800,
  height: 600,
  x: {
    domain: d3.sort(selectedSubset, d => -d.qps).map(d => d.algorithm)
  },
  color: {
    type: "categorical",
    scheme: "tableau10",
    domain: algoDomain
  },
  marginLeft: 90,
  marginTop: 90,
  marginBottom: 50,
  marks: [
    Plot.barY(
      selectedSubset,
      {
        x: "algorithm",
        y: "qps",
        fill: "algorithm"
      }
    ),
    Plot.text(
      selectedSubset,
      {
        x: "algorithm",
        y: "qps",
        text: d => d3.format("d")(d.qps),
        dy: -10
      }
    ),
    Plot.ruleY([0])
  ]
});
```
:::

First and foremost, observe that for the majority of workloads on all datasets we have that one of the two variants of the RD-Index is the best performing data structure.
This is true also when a considerable part of the workload is comprised of range-only or duration-only queries.

On all datasets, when the vast majority of queries are duration-only queries, the best performing data structure is the B-Tree.
The threshold of duration-only queries at which the B-Tree is the best varies from dataset to dataset, and depends also on how many range-duration queries are involved[^why-range-duration].
In particular, for the `Random` the B-Tree has the best performance only when there are at most $\approx 37\%$ range-duration queries and no range-only queries.
On `Flight`, with either $\ge 60\%$  duration-only and no range-only queries, or $76\%$ range-duration and $23\%$  duration-only queries, the B-Tree is the best data structure.
On the other two datasets, the B-Tree outperforms the RD-Index only when the vast majority of queries are range-only.

[^why-range-duration]: The B-Tree indexes only the duration dimension of intervals. Therefore, queries constraining only the range do not benefit from this index, but range-duration queries do, although not as much as duration-only queries.

On `MimicIII`, when almost all the queries are range-only, the strictly best performing data structure is the Grid File. However, as can be seen by the bar chart, the RD-Index performs very similarly.

Finally, observe that the performance of the RD-Index is rather robust to the choice of indexing order: moving the mouse on the ternary plot while observing the bar chart shows that the throughput of RD-Index-dt is similar to the throughput of RD-Index-td.

