---
title: "Indexing Temporal Relations for Range-Duration Queries"
subtitle: "Supplemental material"
author:
  - Matteo Ceccarello <mceccarello@unibz.it>
  - Anton Dignös <dignoes@inf.unibz.it>
  - Johann Gamper <gamper@inf.unibz.it>
  - Christina Khnaisser <christina.khnaisser@usherbrooke.ca>
execute:
  echo: false
format:
  html:
    self-contained: true
---

```{css}
.plot {
  font-size: 12pt;
}
```

```{r setup, echo=F, message=F, warning=F}
library(tidyverse)
library(kableExtra)
library(units)
data <- read_csv("best.csv") %>%
  filter(
    dataset_name %in% c("UZ", "Flight", "Webkit", "MimicIII"),
    time_constraint %in% c("UZ", "UU", "-"),
    duration_constraint %in% c("U", "-")
  ) %>%
  select(
      dataset_name, time_constraint,
      duration_constraint, algorithm_name, qps
  ) %>%
  group_by(dataset_name, time_constraint, duration_constraint) %>%
  distinct(algorithm_name, qps) %>%
  replace_na(list(qps = 0)) %>%
  mutate(
      rank = dense_rank(desc(qps)),
      rank_str = str_c(" {\\footnotesize(", rank, ")}"),
      qps_num = qps,
      qps = qps %>% scales::number(accuracy=1)
  ) %>%
  select(
      dataset = dataset_name,
      time = time_constraint,
      duration = duration_constraint,
      algorithm_name,
      qps, qps_num
  ) %>%
  mutate(
      algorithm_name = case_when(
          algorithm_name == "BTree"          ~ "BTree",
          algorithm_name == "grid-file"      ~ "GridFile",
          algorithm_name == "period-index-*" ~ "Period-Index*",
          algorithm_name == "interval-tree"  ~ "ITree",
          algorithm_name == "rd-index-td"    ~ "RD-Index-td",
          algorithm_name == "rd-index-dt"    ~ "RD-Index-dt",
          algorithm_name == "RTree"          ~ "RTree",
          TRUE ~ algorithm_name
      ),
      algorithm_name = factor(algorithm_name,
          levels = c(
              "RD-Index-td",
              "RD-Index-dt",
              "GridFile",
              "Period-Index*",
              "RTree",
              "ITree",
              "BTree"
          ),
          ordered = TRUE
      )
  ) %>%
  mutate(
      query_type = case_when(
          time == "UZ" && duration == "U" ~ "qrd",
          time == "UU" && duration == "U" ~ "qrd",
          time == "-" && duration == "U"  ~ "qdo",
          time == "UZ" && duration == "-" ~ "qro",
          time == "UU" && duration == "-" ~ "qro",
          T ~ "unknown"
      ),
      query_type = factor(query_type, levels = c("qro", "qdo", "qrd"), ordered = T)
  ) %>%
  ungroup() %>%
  select(
      dataset,
      query = query_type,
      algorithm_name,
      qps, qps_num
  ) %>%
  arrange(algorithm_name)

ojs_define(data = select(data, dataset, query, algorithm_name, qps_num) %>%
  pivot_wider(names_from=query, values_from=qps_num))
```

This document provides supplemental information on the experimental evaluation.

## Mixed-workload queries

The paper reports on the performance of various index structures for uniform workload queries, i.e. range-only queries, duration-only queries, or range-duration queries.

```{r}
#| tbl-cap: "Table 1, as reported in the paper"
#| tbl-cap-location: top
#| column: screen-inset

data %>% 
  select(dataset, query, algorithm_name, qps) %>%
  pivot_wider(
      names_from = "algorithm_name",
      values_from = "qps"
  ) %>%
  mutate(
      dataset = case_when(
          dataset == "UZ" ~ "Random",
          TRUE ~ dataset
      ),
      dataset = factor(dataset,
          levels = c("Random", "Flight", "Webkit", "MimicIII"),
          ordered = TRUE
      )
  ) %>%
  arrange(dataset, query) %>%
  kbl(align="r") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = T, font_size = 14)

```

A natural question is: how does the performance change when these three types of queries are mixed together?
Using data from the above table we can estimate performance figures for mixed workloads.
In particular, the inverse of the throughputs reported in the above table are the average running time of each specific type of query.
Taking the inverse of the weighted average (where weights are the fraction of queries of each type) is the throughput of the mixed workload:
$$
QPS_{mixed} =
\frac{1}{
  f_{ro} / QPS_{ro} +
  f_{do} / QPS_{do} +
  f_{rd} / QPS_{rd}
}
$$
where $f_{ro}$ is the fraction of range only queries in the workload,
$f_{do}$ is the fraction of duration only queries in the workload,
and $f_{rd}$ is the fraction of range-duration queries.

Using the sliders below, we can change the proportion of the three types of queries.

:::{.column-margin}
Setting any of the three values to 1 will give the results also reported in the table above.
:::

```{ojs}
import { aq, op } from '@uwdata/arquero'
viewof range_only_frac = Inputs.range(
  [0, 1],
  {value: 1, step: 0.01, label: "Fraction of range only queries:"}
)
viewof duration_only_frac = Inputs.range(
  [0, 1 - range_only_frac],
  {value: 0, step: 0.01, label: "Fraction of duration only queries:"}
)
range_duration_frac = 1.0 - range_only_frac - duration_only_frac
html`<span>Range-duration queries: ${range_duration_frac.toFixed(2)}</span>`
```

```{ojs}
computed = aq.table(data)
  .params({range_only: range_only_frac, duration_only: duration_only_frac, range_duration: range_duration_frac})
  .derive({
    qps: (d, $) => 1 / ($.range_only / d.qro + $.duration_only / d.qdo + $.range_duration / d.qrd)
  })
bestperf = computed
  .groupby('dataset')
  .rollup({
    maxqps: op.max('qps')
  })
normalized = computed.join(bestperf)
  .derive({
    normqps: d => d.qps / d.maxqps
  })
```

The following plot reports the estimated throughput corresponding to the workload selected above.
The bars lenght is rescaled to the best performance, so that a bar with length 1 corresponds to the best performance and that the plot spans the same domain for all datasets.
The numbers beside each bar report the estimated queries per second.

```{ojs}
//| column: page
Plot.plot({
  grid: false,
  width: 1200,
  height: 1200 / 1.5,
  marginRight: 190,
  marginLeft: 150,
  x: {
    grid: true,
    label: "Queries per second"
  },
  y: {
    grid: false,
    label: "Algorithm"
  },
  facet: {
    data: normalized,
    y: "dataset",
    marginRight: 90
  },
  color: {
    domain: [true, false],
    range: ["orange", "gray"]
  },
  marks: [
    Plot.barX(
      normalized,
      {
        y: "algorithm_name", 
        x: "normqps",
        fill: d => d.algorithm_name == "RD-Index-dt" || d.algorithm_name == "RD-Index-td"
      }
    ),
    Plot.text(
      normalized,
      {
        x: "normqps",
        y: "algorithm_name",
        text: d => d.qps.toFixed(0),
        textAnchor: "start",
        dx: "1em",
        fill: "black"
      }
    )
  ]
})
```

Observe that, as long as there are no duration-only queries in the workload, i.e. when we only have range and range-duration queries,
our index structure `RD-Index` is always the best performing one.

When duration only queries are mixed into the workload, each dataset has a specific threshold of duration only queries such that the throughput of `RD-Index` is better than other index structures.
