\documentclass[manuscript,nonacm,timestamp]{acmart}

\usepackage{todonotes}
\newcommand{\cetodo}[1]{\todo[inline,color=teal!30]{#1}}

\usepackage{tikz}

\newcommand{\BO}[1]{O\left(#1\right)}
\renewcommand{\epsilon}{\varepsilon}

\title{Indexing temporal relations for range-duration queries}

\begin{abstract}
We present a learned index for temporal intervals, supporting range-duration
queries .
\end{abstract}

\begin{document}
\maketitle

\section{Introduction}

\section{Related work}

\section{Preliminaries}

\subsection{Temporal queries}

In this paper we adopt the
\emph{discrete linear time} model~\cite{BohlenDGJ_ebiss2017_temporal-data-management},
where the time domain is an ordered set of discrete \emph{chronons}.
For the continuous case we can simply choose as a chronon an interval of length
equal to the greatest common divisor of the intervals in all relations, and use
it to quantize all the intervals.
A time interval $e = [)$ is a contiguous set of chronons between a starting
time $e_s$ (included) and and ending time $e_e$ (excluded), that is the intervals are \emph{close open}.
Given an interval $e$, we denote with $|e| = e_e - e_s$ its duration, i.e. the number of chronons it spans.
Furthermore, given two intervals $e$ and $e'$ we denote with $e \cap e'$ the set of chronons
shared by the two intervals, which might be empty.
Note that such an intersection is itself an interval, if not empty.

Throughout this paper we will refer to sets of intervals rather than relations with time interval attributes,
for simplicity of exposition.

Our index efficiently supports the three following types of queries (defined as in~\cite{BehrendDGSVRK_SSTD19_period-index}).

\cetodo{Shall we rename it \emph{temporal query?}. Ultimately, also duration queries involve a range of durations, so it might be confusing.}
\begin{definition}[Range query]
    Given a temporal interval $e = [e_s, e_e)$ and a set of intervals $A$,
    a \emph{range query} $Q(A, e)$ is defined as
    \[
        Q(A, e) = \{e' | e' \in A \wedge e' \cap e \ne \emptyset \}
    \]
\end{definition}

\begin{definition}[Duration query]
    Given a duration interval $d = [d_{min}, d_{max}]$ and a set of intervals $A$,
    a \emph{duration query} $Q(A, d)$ is defined as
    \[
        Q(A, d) = \{e | e \in A \wedge |e| \in [d_{min}, d_{max}] \}
    \]
\end{definition}

\begin{definition}[Range-duration query]
    Given a temporal interval $e = [e_s, e_e)$, a duration interval $d = [d_{min}, d_{max}]$,
    and a set of intervals $A$,
    a \emph{range-duration query} $Q(A, e, d)$ is defined as
    \[
        Q(A, e, d) = \{e' | e' \in A \wedge e' \cap e \ne \emptyset \wedge |e'| \in [d_{min}, d_{max}] \}
    \]
\end{definition}

The latter type of query is a combination of the former two.

\section{Data structure}

Our index is a grid, partitioning the intervals into disjoint buckets according
to the durations and the start times of the intervals.
The boundaries between cells are defined by using the empirical cumulative
distribution function of the duration and the starting times, so that
each cell contains the same number of intervals, which corresponds to the
\emph{page size} of our index.
\cetodo{Should we show how to update the data structure?}
Therefore the structure is adaptive to the distribution of the input.

\subsection{Index construction}

Let $s$ be the \emph{page size} parameter.
For an interval $e \in A$, let $CDF_d^A(|e|)$ be the fraction of intervals $e' \in A$
whose duration is
$\le |e|$.
Similarly, let $CDF_s^A(e_s)$ be the fraction of the
intervals whose start time is $\le e_s$.

\cetodo{What happens when there are more than $s^2$ elements in a bucket?
    This happens when there are more than $s^2$ elements with the same duration: they
    all fall in the same bucket.}
\cetodo{We need to describe a way to efficiently compute the cumulative distribution function.}
We partition the intervals in $n / s^2$ buckets: an interval $e$ is assigned to
bucket $\lfloor CDF_d(|e|) \cdot n/2^s \rfloor$.
In particular, we define the $i$-th bucket in the duration dimension as
\[
    A_i^d = \{ e \in A | i = \lfloor CDF_d^A(|e|) \cdot |A|/2^s \rfloor \}
\]
Consider one such bucket $i$, and let $[\bar{d}_{min}, \bar{d}_{max}]$ be the
range of durations spanned by the intervals it contains.
We further partition the intervals in $A_i^d$ according to the distribution
of their starting times:
\[
    A_{i,j}^{d,s} = \{ e \in A_i^d ~|~ j = \lfloor CDF_s(e_s) \cdot |A_i^d|/s \rfloor \}
\]

Our index is a grid built out of the cells $A_{i,j}^{d,s}$.
In each cell, the intervals are sorted by decreasing end time.

We can use a compact representation of the cumulative density function
that is sufficient for indexing into this grid.
Observe that since we always apply a floor operator to the result of the CDF
we can just store the values of the CDF defining the start of a new bucket.

\subsection{Querying the index}

Given a range-duration query $Q(A, e, d)$ with $e=[e_s, e_e)$ and $d=[d_min, d_max]$,
recall that an interval $e'$ satisfies the query if the three following conditions
are met:
\begin{align}
     & |e'| \in [d_{min}, d_{max}] \label{eq:duration-condition} \\
     & e_s < e'_e                  \label{eq:start-condition}    \\
     & e'_s < e_e                  \label{eq:end-condition}
\end{align}

First, find the two indices which identify the range of cells in the duration
dimension each containing at least one interval satisfying
condition~\eqref{eq:duration-condition}.
\begin{align*}
    i_{start} & = \left\lfloor CDF_d^A(d_{min}) \cdot |A| / s^2 \right\rfloor \\
    i_{end}   & = \left\lfloor CDF_d^A(d_{max}) \cdot |A| / s^2 \right\rfloor
\end{align*}
Then, for each $i \in [i_{start}, i_{end}]$, let $[d_{min}^i, d_{max}^i]$
be the range of durations spanned by the $i$-th column.
We consider the cells in the start
dimension in the range $[j_{start}, j_{end}]$ with
\begin{align*}
    j_{start} & = \left\lfloor CDF_{s}^{A_i^d}(e_s - d_{max}^i) \cdot |A_i^d| / s \right\rfloor \\
    j_{end}   & = \left\lfloor CDF_{s}^{A_i^d}(e_e) \cdot |A_i^d| / s \right\rfloor
\end{align*}
In other words,
the start is the first cell such that the upper boundary on the start
time plus the maximum duration in the duration range of the column
is strictly less than the end time of the query
(in Figure~\ref{fig:grid-query} this corresponds to the upper corner of each cell).

For each cell $A_{i,j}^{d,s}$ within the above bounds, we evaluate the query
predicate for all the intervals. Since the intervals are sorted by decreasing
end time, we can stop as soon as we see an interval whose en time is $< e_s$.

Figure~\ref{fig:grid-query} shows an example of a grid together with a query.
Note that each group of durations is partitioned according to start times
in a different way.
As for the query, the dashed lines mark the range of intervals overlapping
the query intervals. Intersecting this area with the diagonals enclosing the
duration part of the query gives the area where query answers lie.

\begin{figure}
    \begin{tikzpicture}[xscale=.2, yscale=.15]
        % Axes
        \draw [<->] (0,60) node (yaxis) [above] {$end$}
        |- (60,0) node (xaxis) [right] {$start$};
        %ticks
        \foreach \x in {0,10,...,50} \draw (\x,10pt) -- (\x,-30pt) node[anchor=north] {\x};
        \foreach \y in {0,10,...,50} \draw (10pt,\y) -- (-30pt,\y) node[anchor=east] {\y};

        % Query point: duration between 8 and 20
        \coordinate (query) at (30,25);
        \draw[thick, red, fill=red!10]
        (17, 25) -- (30, 38) -- (30, 50) -- (5, 25) -- (17, 25) ;
        \fill[red] (query) circle (10pt);
        \draw [red,dashed] (0,25) -- (query);
        \draw [red,dashed] (30,60) -- (query);

        % % Grid structure
        \draw [-] (0,1) -- (60,61);
        \foreach \duration in {5, 15, 27, 50} \draw (0, \duration) -- (60-\duration, 60);
        \foreach \duration in {1, 5, 15, 27, 50} \draw[dashed] (0, \duration) -- (-5, \duration);
        \node[anchor=east] at (-4, 3) {$d\in[1,5)$};
            \node[anchor=east] at (-4, 10) {$d\in[5, 15)$};
        \node[anchor=east] at (-4, 22) {$d\in[15, 27)$};
            \node[anchor=east] at (-4, 38) {$d\in[27, 50)$};

        % grid duration 5
        \foreach \start in {8, 16, 26, 44} \draw (\start, \start+1) -- (\start, \start+5);
        \draw[dashed] (8, 9) -- (8, 5);
        \draw[dashed] (16, 17) -- (16, 5);
        \node[anchor=center] at (12, 5) {$e_s \in [8, 16)$};
        % grid duration 15
        \foreach \start in {7, 15, 22, 33, 40} \draw (\start, \start+5) -- (\start, \start+15);
        % grid duration 27
        \foreach \start in {5, 13, 31} \draw (\start, \start+15) -- (\start, \start+27);
        % grid duration 50
        \draw (5, 5+27) -- (5, 55);
        \draw (20, 20+27) -- (20, 60);

    \end{tikzpicture}
    \caption{Instantiation of the grid over durations and start times, with
        a query. Note that points with the same duration fall on the same diagonal.
        \label{fig:grid-query}}
    \Description{}
\end{figure}

\cetodo{Intervals within a "safe" region are not eve checked, just reported}

\subsection{Analysis}

\begin{theorem}
    Given a set of $n$ intervals, building the index requires time $\BO{n \log n}$.
\end{theorem}
\begin{proof}
    Computing the Empirical Cumulative Distribution Function for the duration measure can be done
    in linear time:
    first find the maximum duration;
    then allocate an array of that many elements;
    then iterate through all the intervals, increasing the count at the appropriate index;
    then iterate through the array, computing the cumulative sum.
    All the above operations can be done in linear time.
    \cetodo{Maybe time linear in the maximum duration. The above arguments assumes that $\max d = \BO{n}$}
    A final scan allows to associate each interval to appropriate bucket.
    These operations can thus be accomplished in time $\BO{n \log n}$.
    Each bucket is then further partitioned according to the start times.
    Let $n_i$ be the size of the $i$-th bucket.
    Then, partitioning it according to the cumulative distribution function
    requires time $\BO{n_i}$, following the same argument as above.
    Therefore, the time for building the second level of the index is
    \[
        \BO{\sum_i n_i \log n_i}
        = \BO{\log n \cdot \sum_i n_i}
        = \BO{n \log n}
    \]
\end{proof}

The following theorem ensures that, by setting the page size $s$ appropriately, the query time
can be made polynomially smaller than the input size.
\begin{theorem}
    Given an index over a set of $n$ intervals, with page size $s$, the time for
    answering a range-duration query is
    \[
        \BO{\frac{n}{s}\log\frac{n}{s} + s^2 + k}
    \]
    where $k$ is the number of intervals matching the query predicate.
    % For $s = \BO{n^\epsilon}$, we have that the query time is
    % \[
    %     \BO{n^\epsilon + k}
    % \]
\end{theorem}
\begin{proof}[Proof draft]
    First, let us establish some notation.
    Let $e = (e_s, e_e)$ be the query interval, and $[d_{min}, d_{max}]$ be the duration range of the query.
    The intervals are arranged into a grid where the \emph{columns} are dictated by the
    durations, and each column is further partitioned in cells according to the start time.
    A cell $C$ is thus identified by a duration range $[d_{min}^C, d_{max}^C)$ and by a start time range
    $[s^C_{min}, s^C_{max})$.
    We deem a column \emph{heavy} if it contains more than $s^2$ intervals, and \emph{light} otherwise.
    Similarly, a cell is \emph{heavy} if it contains more than $s$ intervals, and \emph{light} otherwise.
    A range-duration query defines a \emph{query polygon} in the duration $\times$ start time space.
    For a given query and a cell $C$, we denote by $k_C$ the number of intervals in $C$ matching the query.

    Before proceeding with the proof, we establish some fundamental facts about heavy columns and cells.
    \begin{itemize}
        \item \emph{All intervals in a heavy column have the same duration}.
              If they didn't then we could split the column in two different columns
              with fewer values.
        \item \emph{All intervals in a heavy cell have the same start time}.
              If they didn't then we could split the cell in smaller cells.
        \item \emph{A heavy column does not contain heavy cells}. By the two
              points above a heavy cell in a heavy column will contain intervals
              with the same starting time and duration, i.e. copies of
              the same intervals. Since we are under set semantics this cannot happen.
    \end{itemize}

    The query time is comprised of two parts: first we have to index into the grid to find the correct
    range of cells to query; second we have to filter the intervals in each cell to retrieve
    the ones that are part of the answer.

    As for finding the range of cells, we first have to find the first and last column by duration, which can be
    done by binary search over the $n/s^2$ boundaries between columns in $\BO{n/s^2}$ time.
    There can be $\BO{n / s^2}$ matching columns.
    In each such column, to find the range of cells satisfying the predicate we have to to a binary search
    over the $\BO{n/s}$ intervals in the column (we can have $\BO{n}$ intervals in a heavy column).
    Therefore, the overall time to find the correct cells is
    \[
        \BO{\log\frac{n}{s^2} + \frac{n}{s^2} \log\frac{n}{s}} =
        \BO{\frac{n}{s^2} \log\frac{n}{s}} =
        \BO{\frac{n}{s} \log\frac{n}{s}}
    \]

    Consider now the enumeration of the results from the cells. There are four possible cases for a cell $C$:
    \begin{enumerate}
        \item Both the cell's durations range and start times range fall within the query polygon.
              In such cell, the query constraints on the duration and the start time are satisfied by construction.
              The algorithm considers just the intervals of $C$ which are part of the output (since it iterates over
              the cell's intervals by decreasing end time).
              Therefore, the time spent enumerating intervals from the cell is $\BO{k_C}$.
        \item The cell's duration range satisfies the query constraint, but the starting time range doesn't.
              If the duration colum containing the cell is heavy, then the cell cannot be heavy and the algorithm has to evaluate
              $\BO{s}$ intervals.
              Otherwise the cell might be heavy. In such case, all the start times are equal and satisfy the
              query constraint (otherwise we will not be looking at the cell at all).
              Therefore, the scan of the intervals by decreasing end time visits just the $k_C$ output elements.
        \item The cell's start time satisfies the query constraints, but the duration doesn't.
              In this case we cannot be in a heavy column (which contains a single duration) and thus contains less than $s^2$ intervals.
              If we are in a heavy cell, then all the intervals have the same
              start time:
              \begin{enumerate}
                  \item If the first interval in
                        increasing sorted order of end time (and hence duration)
                        satisfies the duration constraint, scan the intervals by
                        ascending end time, stopping as soon as the duration no longer
                        satisfies the predicate
                  \item Otherwise, do the same in decreasing order of end time
              \end{enumerate}
              In both cases the algorithm will visit just $k_C$ intervals.

              If we are not in a heavy cell, then we need to visit all the $\BO{s}$ intervals.
        \item The cell doesn't completely satisfy neither constraint. In this case neither the column nor the cell can be heavy,
              therefore the algorithm must only evaluate $\BO{s}$ intervals.
    \end{enumerate}
    All the cells of type (1) account for at most $\BO{k}$ in terms of running time.
    As for cells of type (2) there are at most $n/s^2$ of them. Overall they account for $\BO{\frac{n}{s}}$ interval evaluations.
    Cells of type (3) are at most $s$. Therefore, overall they account for $\BO{s^2}$ interval evaluations.
    Finally is then a constant number of cells of type (4) (which are at the corners of the query polygon).
    These cells therefore account for at most $\BO{s}$ interval evaluations in total.
    Overall, the time to evaluate the intervals in the cells is
    \[
        \BO{\frac{n}{s} + s^2 + k}
    \]

    Combining this last result with the time to find the range of cells to query, we have that the overall query time is
    \[
        \BO{\frac{n}{s}\log\frac{n}{s} + s^2 + k}
    \]
\end{proof}

\cetodo{Describe parallelization}

\section{Experimental evaluation}

\bibliographystyle{plain}
\bibliography{references}

\end{document}